### 2021-03-28 final_report/include/CAMT.tex
minor changes
capitalized footnotes

Conversion is based on annotations defining the relation between UD trees and GF ASTs, which we refer to as \textit{dependency configurations}. 
While going into all the notational details is beyond the scope of this project\footnote{a comprehensive description of \texttt{gf-ud}'s annotation scheme is available at \url{github.com/GrammaticalFramework/gf-ud/blob/master/doc/annotations.md}.}, an example of one such configuration is

@@

The problem, which is due to the lack of a root labelled as such and illegal \texttt{ID} and \texttt{HEAD} values, was solved by implementing a simple normalization procedure, now part of the CA library. 
Called every time a UD subtree has to be converted to CoNNL-U, it first replaces the UD label (\texttt{udDEPREL}) of its root node with \texttt{root} and assigns the special value 0 to its \texttt{udHEAD} field. 
After that, it proceeds to ``rescale'' the \texttt{udID} and \texttt{udHEAD} fields of all other nodes so that they end up into the range $[1,n]$, where $n$ is the number of nodes of the extracted subtrees\footnote{see Section \ref{conll} and Figure \ref{types} for definitions of all the fields of UD trees and their CoNNL-U representations.}. \smallskip

@@

Another practical problem that arises when trying to transform a CoNNL-U tree into a GF AST is that \texttt{gf-ud} easily becomes remarkably resource-intensive when dealing with ambiguous, large trees. 
This is due to the fact that conversion in this direction is a nondeterministic search problem \cite{udgf}. 
For this reason, it can be useful to limit the size\footnote{in this context, the size of an alignment is defined as the number of nodes of its largest member.} of the alignments that are actually kept after the extraction (or propagation) procedure. 
Since our CA component works at all levels of abstraction, thus also aligning full clauses, this turned out to be essential and became one of the command-line parameters of the final executables, as mentioned in Appendix \ref{b}. \smallskip

@@

\begin{example}
    ``Won't these short sentences be simpler than that long sentence?''
\end{example}

@@

\begin{itemize}
    \item while GF can parse the SL string in a variety of ways and, as a consequence, return multiple TL linearizations, our program only outputs the first candidate translation, preliminarily discarding trees containing terminal nodes that have no linearization in the TL\footnote{a linearization can be missing, for instance, if a concrete syntax is postprocessed by simply getting rid of an ``illegal'' verb causing compilation problems (cf. Section \ref{grgn})} 
    \item to facilitate the experiments, our standalone has been written so that it can translate several newline-separated sentences at once.
\end{itemize}

@@

To address this problem, the CE module has been modified so that it can try to align individual clauses instead of sentences. 
The idea is to obtain, from each pair of sentence trees to align, a pair of lists of subtrees whose heads are labelled \texttt{root}, \texttt{csubj}, \texttt{ccomp}, \texttt{xcomp}, \texttt{advcl} or  \texttt{acl}\footnote{all these clausal labels are defined in Appendix \ref{a_lab}}.

@@

These lists of ``clauses''\footnote{``clauses'' is in quotes since what we are referring to as a clause includes its subclauses too. This means that the sentence is not segmented into clauses in the literal sense of the term, but rather that one tree per verb is extracted from them.} are then sorted by dependency label and distance from the root, so that in the absence of annotation errors, when there is in fact an evident correspondence, they appear in the same position. Finally, the program tries to align every possible pair of clauses and the pruning procedure described in \ref{pruning} is applied to only keep the ones that are more likely to be correct.

@@

On the one hand, most alignments that do not contain any content word are not useful for the generation of a domain-specific grammar, as many common function words are already covered by the GF RGL\footnote{small alignments containing both function and content words, such as prepositional phrases, are on the other hand extremely useful, as they show which specific function word should be used in a particular context.}. \smallskip

@@

In a preliminary experiment conducted in this way, for instance, the Italian translation ``\textit{il docente discute la Didattica}'' of the English sentence ``\textit{the teacher discusses the didactics}'', was given the minimum score despite being completely correct and even idiomatic\footnote{typically, the adjective ``\textit{didattica}'' is not used in its plural form ``\textit{didattiche}'' when nominalized. The capital D is often used, for this particular term, in official university documents.}. 
Looking at the two reference translations used at the time, it appears clear that such a low score is due to the capitalization of the word ``\textit{Didattica}'' and the arbitrary (but equally correct) lexical choices made by the human translators, who rendered the sentence as ``\textit{il professore discute i metodi didattici}'' and ``\textit{l' insegnante discute la didattica}''\footnote{The terms ``\textit{docente}'', ``\textit{professore}'' and ``\textit{insegnante}'' are almost perfect synonyms, even though ``\textit{docente}'' is mostly used in formal documents to refer to university lecturers, ``\textit{professore}'' is more common in secondary school and colloquial speech, and ``\textit{insegnante}'' is the most generic. As such, it can also refer to elementary school teachers and in some cases even sport coaches. The expression ``\textit{metodi didattici}'' is a perfectly sensible replacement for ``\textit{didattica}''.}. 
Coincidentally (but not unlikely for a language like Italian), this is a case in which these lexical differences affect the choice of the determiners used in the various translations (cf. ``\textit{\underline{l'}insegnante}'' vs. ``\textit{\underline{il} professore}'' and ``\textit{\underline{la} didattica}'' vs. ``\textit{\underline{i} metodi didattici}''), producing a BLEU-1 to 4 score of 0. \smallskip

@@

In Italian, grammar errors often involve contractions such as ``\textit{del}'' (``\textit{di}'' + ``\textit{il}'', in English ``\textit{of the}''), some of which are systematically rendered as two separate words.
Another common case is that of wrong (or at least very confusing) adjective collocation, such as in the translation ``\textit{il libro presenta una tecnica con miglioramenti vari utile}'' (``\textit{the textbook presents a useful technique with various improvements}''), where the adjective ``\textit{utile}'' (``\textit{useful}''), referred to ``\textit{technique}'' (``\textit{tecnica}'') is placed far from such noun, making the sentence hard to interpret\footnote{the manually postprocessed translation is ``\textit{il libro presenta una utile tecnica con miglioramenti vari}''.}.
Grammatical errors in Swedish are less common and, apart from long adjectives never being turned to the comparative degree periphrastically (e.g. ``\textit{relevant\underline{are}}'' instead of ``\textit{mer relevant}''), less systematic. 
Only in one case, for instance, gender is incorrect (``\textit{programbibliotek\underline{en}}''). \smallskip

>>>>

Conversion is based on annotations defining the relation between UD trees and GF ASTs, which we refer to as \textit{dependency configurations}. 
While going into all the notational details is beyond the scope of this project\footnote{A comprehensive description of \texttt{gf-ud}'s annotation scheme is available at \url{github.com/GrammaticalFramework/gf-ud/blob/master/doc/annotations.md}.}, an example of one such configuration is

@@

The problem, which is due to the lack of a root labelled as such and illegal \texttt{ID} and \texttt{HEAD} values, was solved by implementing a simple normalization procedure, now part of the CA library. 
Called every time a UD subtree has to be converted to CoNNL-U, it first replaces the UD label (\texttt{udDEPREL}) of its root node with \texttt{root} and assigns the special value 0 to its \texttt{udHEAD} field. 
After that, it proceeds to ``rescale'' the \texttt{udID} and \texttt{udHEAD} fields of all other nodes so that they end up into the range $[1,n]$, where $n$ is the number of nodes of the extracted subtrees\footnote{See Section \ref{conll} and Figure \ref{types} for definitions of all the fields of UD trees and their CoNNL-U representations.}. \smallskip

@@

Another practical problem that arises when trying to transform a CoNNL-U tree into a GF AST is that \texttt{gf-ud} easily becomes remarkably resource-intensive when dealing with ambiguous, large trees. 
This is due to the fact that conversion in this direction is a nondeterministic search problem \cite{udgf}. 
For this reason, it can be useful to limit the size\footnote{In this context, the size of an alignment is defined as the number of nodes of its largest member.} of the alignments that are actually kept after the extraction (or propagation) procedure. 
Since our CA component works at all levels of abstraction, thus also aligning full clauses, this turned out to be essential and became one of the command-line parameters of the final executables, as mentioned in Appendix \ref{b}. \smallskip

@@

\begin{example}
    ``won't these short sentences be simpler than that long sentence?''
\end{example}

@@

\begin{itemize}
    \item while GF can parse the SL string in a variety of ways and, as a consequence, return multiple TL linearizations, our program only outputs the first candidate translation, preliminarily discarding trees containing terminal nodes that have no linearization in the TL\footnote{A linearization can be missing, for instance, if a concrete syntax is postprocessed by simply getting rid of an ``illegal'' verb causing compilation problems (cf. Section \ref{grgn})} 
    \item to facilitate the experiments, our standalone has been written so that it can translate several newline-separated sentences at once.
\end{itemize}

@@

To address this problem, the CE module has been modified so that it can try to align individual clauses instead of sentences. 
The idea is to obtain, from each pair of sentence trees to align, a pair of lists of subtrees whose heads are labelled \texttt{root}, \texttt{csubj}, \texttt{ccomp}, \texttt{xcomp}, \texttt{advcl} or  \texttt{acl}\footnote{All these clausal labels are defined in Appendix \ref{a_lab}}.

@@

These lists of ``clauses''\footnote{Note that ``clauses'' is in quotes since what we are referring to as a clause includes its subclauses too. This means that the sentence is not segmented into clauses in the literal sense of the term, but rather that one tree per verb is extracted from them.} are then sorted by dependency label and distance from the root, so that in the absence of annotation errors, when there is in fact an evident correspondence, they appear in the same position. Finally, the program tries to align every possible pair of clauses and the pruning procedure described in \ref{pruning} is applied to only keep the ones that are more likely to be correct.

@@

On the one hand, most alignments that do not contain any content word are not useful for the generation of a domain-specific grammar, as many common function words are already covered by the GF RGL\footnote{Small alignments containing both function and content words, such as prepositional phrases, are on the other hand extremely useful, as they show which specific function word should be used in a particular context.}. \smallskip

@@

In a preliminary experiment conducted in this way, for instance, the Italian translation ``\textit{il docente discute la Didattica}'' of the English sentence ``\textit{the teacher discusses the didactics}'', was given the minimum score despite being completely correct and even idiomatic\footnote{Typically, the adjective ``\textit{didattica}'' is not used in its plural form ``\textit{didattiche}'' when nominalized. The capital D is often used, for this particular term, in official university documents.}. 
Looking at the two reference translations used at the time, it appears clear that such a low score is due to the capitalization of the word ``\textit{Didattica}'' and the arbitrary (but equally correct) lexical choices made by the human translators, who rendered the sentence as ``\textit{il professore discute i metodi didattici}'' and ``\textit{l' insegnante discute la didattica}''\footnote{The terms ``\textit{docente}'', ``\textit{professore}'' and ``\textit{insegnante}'' are almost perfect synonyms, even though ``\textit{docente}'' is mostly used in formal documents to refer to university lecturers, ``\textit{professore}'' is more common in secondary school and colloquial speech, and ``\textit{insegnante}'' is the most generic. As such, it can also refer to elementary school teachers and in some cases even sport coaches. The expression ``\textit{metodi didattici}'' is a perfectly sensible replacement for ``\textit{didattica}''.}. 
Coincidentally (but not unlikely for a language like Italian), this is a case in which these lexical differences affect the choice of the determiners used in the various translations (cf. ``\textit{\underline{l'}insegnante}'' vs. ``\textit{\underline{il} professore}'' and ``\textit{\underline{la} didattica}'' vs. ``\textit{\underline{i} metodi didattici}''), producing a BLEU-1 to 4 score of 0. \smallskip

@@

In Italian, grammar errors often involve contractions such as ``\textit{del}'' (``\textit{di}'' + ``\textit{il}'', in English ``\textit{of the}''), some of which are systematically rendered as two separate words.
Another common case is that of wrong (or at least very confusing) adjective collocation, such as in the translation ``\textit{il libro presenta una tecnica con miglioramenti vari utile}'' (``\textit{the textbook presents a useful technique with various improvements}''), where the adjective ``\textit{utile}'' (``\textit{useful}''), referred to ``\textit{technique}'' (``\textit{tecnica}'') is placed far from such noun, making the sentence hard to interpret\footnote{The manually postprocessed translation is ``\textit{il libro presenta una utile tecnica con miglioramenti vari}''.}.
Grammatical errors in Swedish are less common and, apart from long adjectives never being turned to the comparative degree periphrastically (e.g. ``\textit{relevant\underline{are}}'' instead of ``\textit{mer relevant}''), less systematic. 
Only in one case, for instance, gender is incorrect (``\textit{programbibliotek\underline{en}}''). \smallskip

### 2021-03-28 final_report/include/CE.tex
minor layout changes
minor changes
capitalized footnotes

Obviously, the ``full-sentence'' tree pair is itself an alignment, even though a trivial one, as it represents a sentence-level correspondence in a case where, due to the approach being syntactic comparison, the inputs are assumed to be sentences that correspond to each other\footnote{contrary to what it may seem, however, sentence-level alignment is in general not at a trivial task, as it is often the case, especially in certain language pairs, that one \textit{orthographic} sentence, i.e. a sentence defined based on the presence of a full stop, maps to more than one orthographic sentences in the translated text, or vice versa.}.

@@

\subsection{Proposed improvements} \label{improvements}
While one of the most important ideas in the improved version of the CE module proposed in this thesis remains aligning (sub)trees with identical root UD labels, using this criterion alone makes our baseline both unable to detect many of the existing correspondences and prone to extract incorrect ones. This happens both when syntactic similarity is only apparent and when words are aligned based exclusively on the fact that they are the roots of two aligned subtrees\footnote{see Section \ref{heads} for a more detail discussion on head alignment.}.
As a consequence, the objective of this part of the project is to improve both the precision and the recall of the algorithm (or rather, for reasons that will be discussed in Section \ref{metrics}, two approximations of such metrics).

@@

In \cite{divs}, seven classes of divergences are identified: \textit{thematic}, \textit{promotional}, \textit{demotional}, \textit{structural}, \textit{conflational}, \textit{categorial} and \textit{lexical}. 
However, because the author's proposed way to resolve divergences assumes the availability of more than merely syntactic information\footnote{in particular, of a lexicon of Root Lexical Conceptual Structures specifying, for instance, the logical subject, arguments and modifiers of each verb.} and because, working with UD trees, we do not have access to anything but strictly syntactical and morphological annotations, the very task of identifying these distinctions at this level of granularity becomes, in our position, extremely challenging. 
Consequently, we refer to the simpler classification and less formal definitions proposed in \cite{divs2}, where promotional and demotional divergences are merged in a wider-coverage category of \textit{head-swapping} divergences and where lexical divergences, the hardest to handle in the absence of any kind of semantic information, do not appear at all.
In the following, we give definitions and examples for each of these classes of divergences, specifying to what extent and how\footnote{unless otherwise specified, each divergence can be expressed as a set of simple rules in the form of boolean functions taking the pair of UD trees candidate for alignment as input. An example of this can be found in Appendix \ref{cconf}.} each of them is handled in the proposed language-agnostic CE module. We do not attempt to cover all possible cases, but rather provide a few examples as a proof of concept.

@@

\subparagraph*{Conflational divergence} \label{confl}
Conflational divergences involve the translation of two or more words in the source language using a single word that combines their meanings in the target language. A straightforward example is that of compounds, for instance
\begin{example} 
 ``Filippo is interested in \underline{game design}'' VS ``Filippo är intresserad av \underline{spelutvekling}''
\end{example}
but there are also other cases, like
\begin{example} 
 ``I'll come and \underline{say hello} to Bruno and Andrea'' VS ``Verrò a \underline{salutare} Bruno e Andrea''
\end{example} 
where the single word ``\textit{salutare}'' is not a compound but expresses the same meaning as ``\textit{say}'' and ``\textit{hi}'' together. \smallskip
Divergences like the one described in the latter example are usually captured by Criterion \ref{udmatch}, while compounds often need to be taken care of explicitly. For reasons that will become clear later, these cases are discussed in Section \ref{heads}.

@@

These divergences are, while not uncommon, hard to handle without access to a lexicon where entries for verbs are complete with a list of arguments and modifiers\footnote{on the other hand, developing the CA module further could be a way to construct such a ``rich'' lexicon, as will be discussed in Chapter \ref{ch6}.}.

@@

\begin{figure}[H]
 \centering
 \lstinputlisting{pseudocode/Extract.hs}
 \caption[Basic version of the improved CE algorithm]{Basic version of the improved CE algorithm. Instead of aligning full sentence trees and only then extract concepts comparing the structure of the trees and their UD labels only, it recursively extracts pairs of aligned subtrees directly. To do so, for each pair of subtrees occurring in the same context, it tries to apply multiple criteria, implemented as boolean functions telling whether two trees should become the members of an alignment.}
 \label{extr}
\end{figure}

@@

 Without head alignments, the output of the algorithm described so far would be:
 \begin{itemize}
 \item $\langle$``Enrico eats a banana'', ``Enrico mangia una banana''$\rangle$
 \item $\langle$``Enrico'', ``Enrico''$\rangle$
 \item $\langle$``a banana'', ``una banana''$\rangle$
 \item $\langle$``a'', ``una''$\rangle$
 \end{itemize} 
 while introducing head alignment leads to detecting two additional, equally relevant one-word correspondences\footnote{note how, even with head alignments, not all potentially interesting alignments are found. In this example, for instance, it could be useful to align the verb phrase ``\textit{eats a banana}'' with its Italian counterpart ``\textit{mangia una banana}'', whose members are also not, strictly speaking, subtrees of the original sentences. The approach we use for head alignment, as we will discuss in Chapter \ref{ch6}, could be generalized to cases like this.}:
 \begin{itemize}
 \item $\langle$``eats'', ``mangia''$\rangle$
 \item $\langle$``banana'', ``banana''$\rangle$
 \end{itemize} 
\end{example}

@@

In particular, we use a subset of the Parallel UD (PUD) corpus, a set of handcrafted\footnote{for some languages, dependency labels were actually automatically converted to UD format from other standards. Furthermore, manual annotation often only concerns some of fields of the CoNLL-U files, but some manual annotation is generally involved in assigning POS tags and dependency relations.} multilingual treebanks in CoNLL-U format created for the CoNLL 2017 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{mprtud}. This prevents the CE module from failing because of parse errors, even though a small number of annotation inconsistencies and imprecisions is still present. \smallskip

@@

it is hard to deny that ``\textit{boat}'' has been ``translated'' as ``\textit{treno}'', ``\textit{train}'', so our CE module would - and should - be able to detect it, but in most situations it is by no means desirable that such correspondence at the word level is made use of, for instance by storing it in an bilingual lexicon!
As a consequence, each alignment can be marked as:
\begin{itemize}
 \item correct and useful for translation (\texttt{+})\footnote{at least to some extent: we observe that perfect translation equivalents that can be replaced with each other in all contexts are extremely rare, just like perfect synonyms.}
 \item correct but not useful for translation (\texttt{=}). This means that the CE module is working as expected and that we are are faced with a divergence in the broader sense of the term, i.e. potentially due to an idiomatic usage of language
 \item incorrect (\texttt{-}).
\end{itemize} \smallskip

For instance, if we feed the CE module the pair of sentences of Example \ref{extrain}, as a result\footnote{this is what the actual output of the program in evaluation (or ``linearized") mode after annotation looks like. The abbreviations in square brackets refer to alignment Criteria \ref*{udmatch}, \ref{poseq} and \ref{cheads} described in the above.} we get: \smallskip \label{cann}
\begin{verbatim} 
 +he missed the boat|ha perso il treno[UD,POS]1
 +missed|ha perso[UD,POS,HEAD]1
 =the boat|il treno[UD,POS]1
 =boat|treno[UD,POS,HEAD]1
 +the|il[UD]1
\end{verbatim} 
\smallskip

@@

\begin{table}[H]
 \footnotesize
 \centering
 \begin{tabular}{|l|l|l|l|l|}
 \hline
 \textbf{criteria}       & \multicolumn{2}{l|}{\textbf{alignments extracted}} & \multicolumn{2}{l|}{\textbf{correct (+ and =)}} \\ \hline
 \textbf{}       & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}  & \textbf{en-sv}  \\ \hline
 \textbf{matching UD labels}      & 39.39\%  & 37.71\%  & 78.60\%  & 80.24\%  \\ \hline
 \textbf{POS-equivalence}       & 3\%  & 1.82\%  & 83.33\%  & 83.33\%  \\ \hline
 \textbf{known divergence}       & 2.08\%   & 1.36\%  & 52\%   & 55.55\%  \\ \hline
 \textbf{known alignment}       & 0\%  & 0\%  & -  & -  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\end{tabular}}  & 20.95\%  & 24.71\%  & 93.03\%  & 94.76\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + head alignment\end{tabular}}  & 21.78\%  & 19.61\%  & 67.81\%  & 76.74\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}POS-equivalence\\ + head alignment\end{tabular}}  & 1.33\%   & 1.29\%  & 75\%   & 64.70\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + known alignment\end{tabular}}  & 0.25\%   & 0.45\%  & 66.66\%  & 83.33\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\\ + head alignment\end{tabular}} & 10.85\%  & 12.69\%  & 95.38\%  & 90.41\%  \\ \hline
 \end{tabular}
 \caption[Criterion-specific statistics]{Criterion-specific statistics. The leftmost column specifies a criterion or combination of criteria. The central column indicates the percentage of alignments extracted because of each set of criteria, and the rightmost one specifies how many of them were marked as \texttt{+} or \texttt{-}.}
 \label{treas}
 \end{table}

>>>>

Obviously, the ``full-sentence'' tree pair is itself an alignment, even though a trivial one, as it represents a sentence-level correspondence in a case where, due to the approach being syntactic comparison, the inputs are assumed to be sentences that correspond to each other\footnote{Contrary to what it may seem, however, sentence-level alignment is in general not at a trivial task, as it is often the case, especially in certain language pairs, that one \textit{orthographic} sentence, i.e. a sentence defined based on the presence of a full stop, maps to more than one orthographic sentences in the translated text, or vice versa.}.

@@

\subsection{Proposed improvements} \label{improvements}
While one of the most important ideas in the improved version of the CE module proposed in this thesis remains aligning (sub)trees with identical root UD labels, using this criterion alone makes our baseline both unable to detect many of the existing correspondences and prone to extract incorrect ones. This happens both when syntactic similarity is only apparent and when words are aligned based exclusively on the fact that they are the roots of two aligned subtrees\footnote{See Section \ref{heads} for a more detail discussion on head alignment.}.
As a consequence, the objective of this part of the project is to improve both the precision and the recall of the algorithm (or rather, for reasons that will be discussed in Section \ref{metrics}, two approximations of such metrics).

@@

In \cite{divs}, seven classes of divergences are identified: \textit{thematic}, \textit{promotional}, \textit{demotional}, \textit{structural}, \textit{conflational}, \textit{categorial} and \textit{lexical}. 
However, because the author's proposed way to resolve divergences assumes the availability of more than merely syntactic information\footnote{In particular, of a lexicon of Root Lexical Conceptual Structures specifying, for instance, the logical subject, arguments and modifiers of each verb.} and because, working with UD trees, we do not have access to anything but strictly syntactical and morphological annotations, the very task of identifying these distinctions at this level of granularity becomes, in our position, extremely challenging. 
Consequently, we refer to the simpler classification and less formal definitions proposed in \cite{divs2}, where promotional and demotional divergences are merged in a wider-coverage category of \textit{head-swapping} divergences and where lexical divergences, the hardest to handle in the absence of any kind of semantic information, do not appear at all.
In the following, we give definitions and examples for each of these classes of divergences, specifying to what extent and how\footnote{Unless otherwise specified, each divergence can be expressed as a set of simple rules in the form of boolean functions taking the pair of UD trees candidate for alignment as input. An example of this can be found in Appendix \ref{cconf}.} each of them is handled in the proposed language-agnostic CE module. We do not attempt to cover all possible cases, but rather provide a few examples as a proof of concept.

@@

\subparagraph*{Conflational divergence} \label{confl}
Conflational divergences involve the translation of two or more words in the source language using a single word that combines their meanings in the target language. A straightforward example is that of compounds, for instance
\begin{example} 
 ``Filippo is interested in \underline{game development}'' VS ``Filippo är intresserad av \underline{spelutveckling}''
\end{example}
but there are also other cases, like
\begin{example} 
 ``I'll come and \underline{say hello} to Bruno and Andrea'' VS ``Verrò a \underline{salutare} Bruno e Andrea''
\end{example} 
where the single word ``\textit{salutare}'' is not a compound but expresses the same meaning as ``\textit{say}'' and ``\textit{hi}'' together. \smallskip
Divergences like the one described in the latter example are usually captured by Criterion \ref{udmatch}, while compounds often need to be taken care of explicitly. For reasons that will become clear later, these cases are discussed in Section \ref{heads}.

@@

These divergences are, while not uncommon, hard to handle without access to a lexicon where entries for verbs are complete with a list of arguments and modifiers\footnote{On the other hand, developing the CA module further could be a way to construct such a ``rich'' lexicon, as will be discussed in Chapter \ref{ch6}.}.

@@

\begin{figure}[h]
 \centering
 \lstinputlisting{pseudocode/Extract.hs}
 \caption[Basic version of the improved CE algorithm]{Basic version of the improved CE algorithm. Instead of aligning full sentence trees and only then extract concepts comparing the structure of the trees and their UD labels only, it recursively extracts pairs of aligned subtrees directly. To do so, for each pair of subtrees occurring in the same context, it tries to apply multiple criteria, implemented as boolean functions telling whether two trees should become the members of an alignment.}
 \label{extr}
\end{figure}

@@

 Without head alignments, the output of the algorithm described so far would be:
 \begin{itemize}
 \item $\langle$``Enrico eats a banana'', ``Enrico mangia una banana''$\rangle$
 \item $\langle$``Enrico'', ``Enrico''$\rangle$
 \item $\langle$``a banana'', ``una banana''$\rangle$
 \item $\langle$``a'', ``una''$\rangle$
 \end{itemize} 
 while introducing head alignment leads to detecting two additional, equally relevant one-word correspondences\footnote{Note how, even with head alignments, not all potentially interesting alignments are found. In this example, for instance, it could be useful to align the verb phrase ``\textit{eats a banana}'' with its Italian counterpart ``\textit{mangia una banana}'', whose members are also not, strictly speaking, subtrees of the original sentences. The approach we use for head alignment, as we will discuss in Chapter \ref{ch6}, could be generalized to cases like this.}:
 \begin{itemize}
 \item $\langle$``eats'', ``mangia''$\rangle$
 \item $\langle$``banana'', ``banana''$\rangle$
 \end{itemize} 
\end{example}

@@

In particular, we use a subset of the Parallel UD (PUD) corpus, a set of handcrafted\footnote{For some languages, dependency labels were actually automatically converted to UD format from other standards. Furthermore, manual annotation often only concerns some of fields of the CoNLL-U files, but some manual annotation is generally involved in assigning POS tags and dependency relations.} multilingual treebanks in CoNLL-U format created for the CoNLL 2017 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{mprtud}. This prevents the CE module from failing because of parse errors, even though a small number of annotation inconsistencies and imprecisions is still present. \smallskip

@@

it is hard to deny that ``\textit{boat}'' has been ``translated'' as ``\textit{treno}'', ``\textit{train}'', so our CE module would - and should - be able to detect it, but in most situations it is by no means desirable that such correspondence at the word level is made use of, for instance by storing it in an bilingual lexicon!
As a consequence, each alignment can be marked as:
\begin{itemize}
 \item correct and useful for translation (\texttt{+})\footnote{At least to some extent: we observe that perfect translation equivalents that can be replaced with each other in all contexts are extremely rare, just like perfect synonyms.}
 \item correct but not useful for translation (\texttt{=}). This means that the CE module is working as expected and that we are are faced with a divergence in the broader sense of the term, i.e. potentially due to an idiomatic usage of language
 \item incorrect (\texttt{-}).
\end{itemize} \smallskip

For instance, if we feed the CE module the pair of sentences of Example \ref{extrain}, as a result\footnote{This is what the actual output of the program in evaluation (or ``linearized") mode after annotation looks like. The abbreviations in square brackets refer to alignment Criteria \ref*{udmatch}, \ref{poseq} and \ref{cheads} described in the above.} we get: \smallskip \label{cann}
\begin{verbatim} 
 +he missed the boat|ha perso il treno[UD,POS]1
 +missed|ha perso[UD,POS,HEAD]1
 =the boat|il treno[UD,POS]1
 =boat|treno[UD,POS,HEAD]1
 +the|il[UD]1
\end{verbatim} 
\smallskip

@@

\begin{table}[h]
 \footnotesize
 \centering
 \begin{tabular}{|l|l|l|l|l|}
 \hline
 \textbf{criteria}       & \multicolumn{2}{l|}{\textbf{alignments extracted}} & \multicolumn{2}{l|}{\textbf{correct (+ and =)}} \\ \hline
 \textbf{}       & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}  & \textbf{en-sv}  \\ \hline
 \textbf{matching UD labels}      & 39.39\%  & 37.71\%  & 78.60\%  & 80.24\%  \\ \hline
 \textbf{POS-equivalence}       & 3\%  & 1.82\%  & 83.33\%  & 83.33\%  \\ \hline
 \textbf{known divergence}       & 2.08\%   & 1.36\%  & 52\%   & 55.55\%  \\ \hline
 \textbf{known alignment}       & 0\%  & 0\%  & -  & -  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\end{tabular}}  & 20.95\%  & 24.71\%  & 93.03\%  & 94.76\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + head alignment\end{tabular}}  & 21.78\%  & 19.61\%  & 67.81\%  & 76.74\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}POS-equivalence\\ + head alignment\end{tabular}}  & 1.33\%   & 1.29\%  & 75\%   & 64.70\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + known alignment\end{tabular}}  & 0.25\%   & 0.45\%  & 66.66\%  & 83.33\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\\ + head alignment\end{tabular}} & 10.85\%  & 12.69\%  & 95.38\%  & 90.41\%  \\ \hline
 \end{tabular}
 \caption[Criterion-specific statistics]{Criterion-specific statistics. The leftmost column specifies a criterion or combination of criteria. The central column indicates the percentage of alignments extracted because of each set of criteria, and the rightmost one specifies how many of them were marked as \texttt{+} or \texttt{-}.}
 \label{treas}
 \end{table}

### 2021-03-28 final_report/include/CP.tex
capitalized footnotes

There are two scenarios in which CP can prove useful:\smallskip
\begin{enumerate}
    \item when working on a parallel text $A$ in more than two languages. In this case, CP can be used, after CE has been performed on a pair of languages $\langle L_1,L_2 \rangle$, to propagate the extracted alignments in a third version of the text in a new language $L_3$. In this case, the expectation is for the program to be able to propagate the vast majority\footnote{even though handling common translation divergences help, free translation involving radical differences in sentence structure can make it so that some concepts lack a clear counterpart in the text in $L_3$.} of the concepts in the set
    \item when working on two bilingual parallel texts $A$ and $B$ with one language in common, say $A$ in two languages $\langle L_1,L_2 \rangle$ and $B$ in two languages $\langle L_2, L_3 \rangle$. After performing CE on the first pair, CP can be used to look for the extracted concepts in the $L_2$ version of the second and, if they do appear in such text, find their counterparts in $L_3$. Of course, the number of concepts that can be propagated in this scenario can be significantly lower, especially if the two parallel texts do not belong to the same domain.
\end{enumerate}\smallskip

@@

Our algorithm is meant to be as general as possible, and as such it is not optimized for the first scenario, but rather covers both without making any assumptions in terms of sentence alignment\footnote{a way to make the algorithm more efficient for the first scenario could in fact be to rely on the fact that the $n$-lingual parallel text is sentence-aligned, keep track of the sentence from which a particular concept has been extracted and only try to find a equivalent in the corresponding $L_3$ sentence.}. 
It works as follows: for each alignment obtained by performing CE on text $A$, we consider the dependency tree in language $L_2$. We then look for such tree among all subtrees the $L_2$ version of text $B$ is composed of. If a sentence in the $L_2$ version of text $B$ does contain $a_i^2$, we align such sentence to its $L_3$ counterpart using the same procedure as described in Section \ref{extrain} for CE. Finally, if multiple possible correspondences are found, the $L_3$ dependency subtree which is closest in depth to $a_i^2$ is selected. This is meant to avoid, for instance, mistakenly aligning a full sentence tree to a dependency tree extracted when aligning heads, which would otherwise be likely, as their heads are obviously both labelled \texttt{root}. In the next two sections, we discuss some important details of the algorithm, and in Section \ref{heads2} in particular will get to other issues related to head alignment.

>>>>

There are two scenarios in which CP can prove useful:\smallskip
\begin{enumerate}
    \item when working on a parallel text $A$ in more than two languages. In this case, CP can be used, after CE has been performed on a pair of languages $\langle L_1,L_2 \rangle$, to propagate the extracted alignments in a third version of the text in a new language $L_3$. In this case, the expectation is for the program to be able to propagate the vast majority\footnote{Even though handling common translation divergences help, free translation involving radical differences in sentence structure can make it so that some concepts lack a clear counterpart in the text in $L_3$.} of the concepts in the set
    \item when working on two bilingual parallel texts $A$ and $B$ with one language in common, say $A$ in two languages $\langle L_1,L_2 \rangle$ and $B$ in two languages $\langle L_2, L_3 \rangle$. After performing CE on the first pair, CP can be used to look for the extracted concepts in the $L_2$ version of the second and, if they do appear in such text, find their counterparts in $L_3$. Of course, the number of concepts that can be propagated in this scenario can be significantly lower, especially if the two parallel texts do not belong to the same domain.
\end{enumerate}\smallskip

@@

Our algorithm is meant to be as general as possible, and as such it is not optimized for the first scenario, but rather covers both without making any assumptions in terms of sentence alignment\footnote{A way to make the algorithm more efficient for the first scenario could in fact be to rely on the fact that the $n$-lingual parallel text is sentence-aligned, keep track of the sentence from which a particular concept has been extracted and only try to find a equivalent in the corresponding $L_3$ sentence.}. 
It works as follows: for each alignment obtained by performing CE on text $A$, we consider the dependency tree in language $L_2$. We then look for such tree among all subtrees the $L_2$ version of text $B$ is composed of. If a sentence in the $L_2$ version of text $B$ does contain $a_i^2$, we align such sentence to its $L_3$ counterpart using the same procedure as described in Section \ref{extrain} for CE. Finally, if multiple possible correspondences are found, the $L_3$ dependency subtree which is closest in depth to $a_i^2$ is selected. This is meant to avoid, for instance, mistakenly aligning a full sentence tree to a dependency tree extracted when aligning heads, which would otherwise be likely, as their heads are obviously both labelled \texttt{root}. In the next two sections, we discuss some important details of the algorithm, and in Section \ref{heads2} in particular will get to other issues related to head alignment.
